---
title: "ENVS193DS final"
author: Sophie Pardehpoosh
date: "2026-03-13"
format:
  html:
    toc: true # use this to display a table of contents
execute:
  message: false # make sure messages don't show up
  warning: false # make sure warnings don't show up
---

## [Github repository](https://github.com/spardehpoosh/ENVS-193DS_winter-2026_final.git)

# Set up

```{r}
#| label: reading in packages and data
#| message: false

library(tidyverse) # general use
library(janitor) # cleaning data frames
library(here) # file/folder organization
library(readxl) # reading .xlsx files

# reading in data
kelp <- read_csv(here("data", "kelp_data.csv")) # storing kelp data as an object called kelp for problem 2
```


# Problem 1. Research Writing 

## a. Transparent statistical methods

In part 1, my coworker used simple linear regression to determine whether or not precipitation predicts flooded wetland area. It is "simple" since there is one predictor. 

In part 2, they used a Kruskal-Wallis test to determine whether medians differed accross water year clasification. Since they were comparing more than two predictors they would use ANOVA but since they are comparing medians rather than means, they need the non parametric version which is Kruskal Wallis. 


## b. Figure needed

I would suggest that my coworkers includes a box plot comparing median flooded wetland area across the five water year classifications. The y-axis would have the flooded wetland area (m^2^), while the x axis would label the five water classifications. There would be five boxes on the plot. Each of the five would have a line for the median, with upper and lower quartiles at the top and bottom of the box. Individual data points would be layed on top of the box plots to visualize the raw data. 

## c. Suggestions for rewriting

Flooded wetland area (m^2^) did not change with precipitation (linear regression, F(df, df) = test statistic, R^2^ = coefficient of determination, p = 0.11, $\alpha$ = significance level). 
Median flooded wetland area (m^2^) did not differ among water year clasifications  (wet, above normal, below normal, dry, critical drought) (Kruskal-Wallis test, $\chi$^2^(df) = test statistic, p = 0.12, $\alpha$ = significance level).


## d. interpretation

Wetland management type is a nominnal categorical variable that could influence flooded area of a wetland. In the paper, they state that local wetland management practices impact wetland flooding, and that "semi-permanent wetlands" is a management type that would increase flooded area during parts of the year. 


# Problem 2. Data visualization

## a. Cleaning and summarizing

```{r}

kelp_clean <- kelp |>  # creating kelp_clean data set from kelp data
  clean_names() |> # cleaning the column names so only lower case letters and underscores are used
  filter(site %in% c("IVEE", "NAPL", "CARP")) |># filter to only include these 3 sites
  mutate(site_name = case_when( # creating a new column called site_name and relabelling the three sites with full names
  site == "NAPL" ~ "Naples",
  site == "IVEE" ~ "Isla Vista",
  site == "CARP" ~ "Carpinteria"
)) |> 
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |> # filtering to only include observations from 2000-09-22 date at transect 6 and quadrat 40 
  mutate(site_name = as_factor(site_name), # making sure that categorical variables are factors
       site_name = fct_relevel(site_name, # order of factor levels
                               "Naples", "Isla Vista", "Carpinteria")) |> 
  group_by(site_name, year) |> # group data by site and year so that summaries are calculated in these groups
  summarize(mean_fronds = mean(fronds)) |> # calculate mean number of fronds in each site/year group
  ungroup() # ungrouping for proceeding operations

slice_sample(kelp_clean, n=5) # displaying 5 random rows of the summarized and cleaned data frame 

```






