---
title: "ENVS193DS final"
author: Sophie Pardehpoosh
date: "2026-03-13"
format:
  html:
    toc: true # use this to display a table of contents
execute:
  message: false # make sure messages don't show up
  warning: false # make sure warnings don't show up
---

## [Github repository](https://github.com/spardehpoosh/ENVS-193DS_winter-2026_final.git)

# Set up

```{r}
#| label: reading in packages and data
#| message: false

library(tidyverse) # general use
library(janitor) # cleaning data frames
library(here) # file/folder organization
library(readxl) # reading .xlsx files

# reading in data
kelp <- read_csv(here("data", "kelp_data.csv")) # storing kelp data as an object called kelp for problem 2
mopl <- read_csv(here("data", "MOPL_nest-site_selection_Duchardt_et_al._2019.csv")) # storing data for problem 3 as an object called mopl
```


# Problem 1. Research Writing 

## a. Transparent statistical methods

In part 1, my coworker used simple linear regression to determine whether or not precipitation predicts flooded wetland area. It is "simple" since there is one predictor. 

In part 2, they used a Kruskal-Wallis test to determine whether medians differed accross water year clasification. Since they were comparing more than two predictors they would use ANOVA but since they are comparing medians rather than means, they need the non parametric version which is Kruskal Wallis. 


## b. Figure needed

I would suggest that my coworkers includes a box plot comparing median flooded wetland area across the five water year classifications. The y-axis would have the flooded wetland area (m^2^), while the x axis would label the five water classifications. There would be five boxes on the plot. Each of the five would have a line for the median, with upper and lower quartiles at the top and bottom of the box. Individual data points would be layed on top of the box plots to visualize the raw data. 

## c. Suggestions for rewriting

Flooded wetland area (m^2^) did not change with precipitation (linear regression, F(df, df) = test statistic, R^2^ = coefficient of determination, p = 0.11, $\alpha$ = significance level). 
Median flooded wetland area (m^2^) did not differ among water year clasifications  (wet, above normal, below normal, dry, critical drought) (Kruskal-Wallis test, $\chi$^2^(df) = test statistic, p = 0.12, $\alpha$ = significance level).


## d. Interpretation

Wetland management type is a nominnal categorical variable that could influence flooded area of a wetland. In the paper, they state that local wetland management practices impact wetland flooding, and that "semi-permanent wetlands" is a management type that would increase flooded area during parts of the year. 


# Problem 2. Data visualization

## a. Cleaning and summarizing

```{r}
#| label: cleaning and summarizing

kelp_clean <- kelp |>  # creating kelp_clean data set from kelp data
  clean_names() |> # cleaning the column names so only lower case letters and underscores are used
  filter(site %in% c("IVEE", "NAPL", "CARP")) |># filter to only include these 3 sites
  mutate(site_name = case_when( # creating a new column called site_name and relabelling the three sites with full names
  site == "NAPL" ~ "Naples",
  site == "IVEE" ~ "Isla Vista",
  site == "CARP" ~ "Carpinteria"
)) |> 
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |> # filtering out frond observations from date 2000-09-22 at transect 6 and quadrat 40 
  mutate(site_name = as_factor(site_name), # making sure that categorical variables are factors
       site_name = fct_relevel(site_name, # order of factor levels
                               "Naples", "Isla Vista", "Carpinteria")) |> 
  group_by(site_name, year) |> # group data by site and year so that summaries are calculated in these groups
  summarize(mean_fronds = mean(fronds)) |> # calculate mean number of fronds in each site/year group
  ungroup() # ungrouping for proceeding operations

slice_sample(kelp_clean, n=5) # displaying 5 random rows of the summarized and cleaned data frame 

```

## b. Understanding the data

We filtered out frond observations from the date 2000-09-22 in transect 6 in quad 40 because there is a missing value for fronds. In the column for "fronds" for that date, transect and quad, the value is -99999. In the "methods and protols" section of the dataset page, they state "The count values were given a NA (-99999) if: (1) data were not collected for a particular species or (2) a datasheet was lost" 

## c. Visualize the data

```{r}
#| label: fronds figure

# base layer: ggplot
ggplot(data = kelp_clean, # making figure from kelp_clean data set
       aes(x = year, # having date on the x-axis
           y = mean_fronds, # mean fronds on the y-axis
           color = site_name)) + # different colors for each site 
  # first layer: points
  geom_point() +
  # second layer: lines 
  geom_line() + 
  labs(x = "Year", # relabelling the x-axis
       y = "Mean kelp fronds per year", # relabelling the y-axis
       title = "Sites vary in mean kelp fronds per year", # adding a title to the figure 
       color = "Site") + # capitalizing the word "site" in the key 
  theme_minimal() + # theme for grid with no border
  theme(legend.position = "right", # putting the legend on the right
        panel.background = element_blank(), # getting rid of panel background
        plot.title = element_text(hjust = 0.5), # centering title
        text = element_text(family = "Times")) + # changing font
   scale_color_manual( # changing the color of each site line
    values = c("Isla Vista" = "tan", # changing color of line for Isla Vista 
               "Naples" = "aquamarine3", #changing color of line for Naples
               "Carpinteria" = "darkgrey"))  # changing color of line for Santa Cruz - Diablo
```

## d. Interpretation

According to the figure, Isla Vista is the site that has the most variable mean kelp fronds per year, while Carpinteria seems to have the least variable mean kelp fronds per year. 

# Problem 3. Data Analysis

```{r}
#| label: exploring data
#| include: false

mopl_clean <- mopl |> # starting with mopl data
  # making sure id is seen as a factor 
  clean_names() |> # clean column names
  mutate(id = as_factor(id),
         id = fct_relevel(id, 
                          "used",
                          "available")) |> 
   # select to only include columns of interest 
  select(id, vo_nest, edgedist)

```

```{r}
#| label: checking structure
#| include: false

str(mopl_clean)
```

```{r}
#| label: displaying random rows in data frame
#| include: false

slice_sample(
  mopl_clean,
  n = 10)

```


## a. Response variable

0s represent available locations where nest weren't observed, while 1s represent available locations where snowy plovers actually nested. 

## b. Purpose of study













